Detailed structure of phospholipid bilayers with negatively charged lipids and salt concentrations from molecular dynamics simulations including the effects of electronic polarization

### 2019-05-13 Manuscript/manuscript.tex
Improvements in the Methods section (small)

\begin{table*}[tbp]
\centering
\caption{  List of MD simulations reporting their respective
           lipid composition,
           used counterions,
           added buffer concentrations,
           amounts of individual molecules,
           and the deployed models.
           The fatty acid chains for all lipids are plamitoyl (sn-1) and oleoyl (sn-2).
         }\label{tbl:sim-list}
\begin{tabular}{l l | r r r r | r r | c c }
   \multicolumn{2}{c}{ }  &  \multicolumn{4}{c}{added buffer conc. / mM}    & \multicolumn{2}{c}{no. molecules} &  \multicolumn{2}{c}{simulated with model}  \\
PC:PS ratio & counterions &  \ce{K^+}  &  \ce{Na^+} & \ce{Ca^{2+}} & \ce{Cl^-}      & \ce{H2O} &  PC:PS                 &  ECC-lipids  &  Lipid17    \\
  \hline
0:1 & \ce{K^+}   &      0  &      0  &      0  &      0  &  3600  &  0:72  &  X  &  X  \\ 
0:1 & \ce{Na^+}  &      0  &      0  &      0  &      0  &  3600  &  0:72  &  X  &  X  \\ 
  \hline
10:1 & \ce{K^+}  &      0  &      0  &      0  &      0  &  3600  &  60:6  &  X  &  -  \\ 
10:1 & \ce{Na^+} &      0  &      0  &      0  &      0  &  3600  &  60:6  &  X  &  -  \\ 
  \hline
1:1 & \ce{K^+}   &      0  &      0  &      0  &      0  &  5253  &  64:64  &  X  &  X  \\ 
1:1 & \ce{Na^+}  &      0  &      0  &      0  &      0  &  5253  &  64:64  &  X  &  X  \\ 
  \hline
4:1 & \ce{K^+}   &      0  &      0  &      0  &      0  &  3600  &  48:12  &  X  &  -  \\ 
4:1 & \ce{Na^+}  &      0  &      0  &      0  &      0  &  3600  &  48:12  &  X  &  -  \\ 
  \hline
5:1 & \ce{K^+}   &      0  &      0  &      0  &      0  &  3600  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &      0  &      0  &      0  &      0  &  3600  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &      0  &      0  &     78  &     78  &  3561  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &      0  &      0  &    125  &    125  &  3561  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &      0  &      0  &    202  &    202  &  3561  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &      0  &      0  &    409  &    409  &  3522  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &      0  &      0  &    621  &    621  &  3483  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &      0  &    621  &      0  &    621  &  3483  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &      0  &   1510  &      0  &   1510  &  3377  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &      0  &   3002  &      0  &   3002  &  3213  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &    621  &      0  &      0  &    621  &  3483  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &   1510  &      0  &      0  &   1510  &  3377  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &   3002  &      0  &      0  &   3002  &  3213  &  60:12  &  X  &  X  \\ 
  \hline
\end{tabular}
\end{table*}

@@

In the reference simulations with the standard Amber force field,
the Lipid14 parameters~\cite{dickson14} for POPC, the Lipid17 parameters for POPS \cite{lipid17-future},
the TIP3P water model~\cite{jorgensen83}, and ion models by Dang and coworkers~\cite{smith94,chang1999,dang2006} were used.
The previously generated \cite{botan15} Lipid14 parameters for POPC in Gromacs format were downloaded from Ref.~\citenum{lipid14files}. 
The Lipid17 parameters for POPS were obtained from AmberTools18 \cite{amber18} 
and converted to Gromacs format using acpype \cite{acpype}.  
The ion model by Dang and coworkers~\cite{smith94,chang1999,dang2006} was used with Amber lipids because
the default Amber ion parameters \cite{aqvist90} led to the artificial clustering of ions in solution \cite{NMRlipidsIV}.
In the ECC-lipid simulations,
the ECC-POPC parameters~\cite{melcr18} (available from Ref. \citenum{ECC-POPC_nacl_cacl2_files}), 
the ECC-POPS parameteres derived in this work (see above), 
the SPC/E~\cite{Berendsen1987} water model, and ECC-ions \cite{martinek17, kohagen16, Pluharova2014} were used.
The SPC/E water model was selected because its lower dielectric constant is consistent with the
ECC concept \cite{leontyev11,leontyev14}.
\begin{table}[tbp]
  \caption{Simulation parameters}
  \label{tbl:mdpar}
  \begin{tabular}{ll}
    simulation property & parameter   \\
    \hline
    time-step           & 2~fs         \\
    equilibration time  & 50~ns  \\
    total simulation time     & 1~$\mu$s  \\
    temperature         & 298~K       \\
    thermostat          & v-rescale  \cite{bussi07}   \\
    barostat            & Parrinello-Rahman, semi-isotropic \cite{parrinello81} \\
    long-range electrostatics & PME  \cite{darden93}  \\
    cut-off scheme      & Verlet \cite{Pall13}      \\
    Coulomb and VdW cut-off & 1.0~nm \\
    constraints         & LINCS, only hydrogen atoms \cite{hess97} \\
    constraints for water & SETTLE  \cite{miyamoto92} \\
    \hline
  \end{tabular}
\end{table}

The C--H bond order parameters were calculated directly from Eq. \ref{OPequation}.
Average over the time for individual lipids was calculated first and
the standard error of the mean of the average over lipids was then used as the
error estimate \cite{botan15,ollila16,NMRlipidsIV}.
Python program ({\tt scripts/calcOrderParameters.py}) that uses the
MDAnalysis library \cite{agrawal11,gowers16} is available in Ref. \citenum{MATCHgit}. 
The ion number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.
The salt concentrations in buffer before solvating the lipids, reported in the used experimental data set \cite{roux90},
were calculated as $[{\rm salt}]=N_{\rm c} \times[{\rm water}]\,/\,N_{\rm w}$,
where $N_{\rm c}$ is the number of cations in simulation, $[{\rm water}]$\,=\,55.5~M and $N_{\rm w}$
is the number of water molecules in the simulation.
As discussed in our previous work \cite{NMRlipidsIV}, the hydration levels of
multilamellae are expected to be sufficiently similar in the used simulations and reference experiments \cite{roux90}.

%The simulated system consisted of lipids (POPC and POPS) forming a single lipid bilayer,
%an amount of water molecules comparable to the hydration levels in experiments, 
%\todo{Check this claim. SAMULI: Check discussin in the NMRlipids IV manuscript}
%\todo{EDIT: Taken from NMRlipids4 as an interesting information, to be edited.}
%In the experimental $^2$H NMR literature data used in this work~\cite{browning80,roux90},
%the lipids were first soluted to the buffer and then centrifuged to a pellet which
%was used in the measurements. Such samples have lower lipid concentration
%(approximately 10~wt~\% of lipids~\cite{browning80,roux88,roux90}) than 
%gravimetric samples (60~wt~\%) and simulations (approximately 50-60~wt~\%) in this work.
%Larger multilamellar repeat distances are expected in the samples with lower lipid
%concentrations due to the swelling caused by electrostatic repulsion in pure PS lipid systems~\cite{millman82}.
%However, the PS headgroup order parameters measured from gravimetric sample in this work
%are in good agreement with the results from centrifuged sample in the literature \cite{browning80} (Fig. \ref{HGorderParameters}).
%Furthermore, the equlibrium repeat distance rapidly decreases with the addtion of monovalent
%salts and is close to the simulation box sizes already above 500~mM concentrations~\cite{millman82,rand89}.
%Therefore, the hydration levels of multilamellae are expected to be sufficiently similar
%in the used simulations and reference experiments.

@@

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[height=\figheight]{../img/ecc_pops/density_profiles_na-k-counterions_wat_phos_compar_purePOPS_ecclipids-lipid17.pdf}
  \includegraphics[height=\figheight]{../img/ecc_pops/density_profiles_na-k-counterions_wat_phos_compar_5PC-1PS_ecclipids-lipid17.pdf}
  \caption{\label{fig:POPS-counterions-dens}
    Number density profiles of \ce{K^{+}} and \ce{Na^{+}} counterions along the membrane normal axis
    in ECC-lipid (solid lines) and Lipid17/Dang (dashed lines) simulations of pure POPS (left) or POPC:POPS (5:1, right) bilayers.  
    The density profiles of phosphate groups and water are divided by 4 and 100, respectively.  
}
\end{figure} 
Binding affinity of monovalent ions to lipid bilayers depends strongly on
force field parameters \cite{catte16,NMRlipidsIV},
which is the case also here as the sodium counterion binding affinity to
POPS lipid bilayer is stronger in Lipid17/ Dang simulation than in
ECC-POPS simulation (Fig. \ref{fig:POPS-counterions-dens}).
The cation binding affinities to lipid bilayers in simulations have been previously evaluated
against experiments using the order parameters of $\alpha$ and $\beta$ carbons in the
phosphatidylcholine (PC) lipid headgroup (see Fig.~\ref{fig:chemstruct_pc_ps} for the labeling) \cite{catte16,melcr18,NMRlipidsIV},
which decrease proportionally to the bound charge according to the ''electrometer concept'' \citep{seelig87}.
Althought the response of PS lipid headgroup order parameters to the bound charge is also systematic,
it is less well undestood and the addition of cations may cause phase transitions in negatively charged
bilayers \cite{feigenson86,mattai89,roux91,roux90}. Therefore, the cation binding affinity to bilayers with
PS lipids has been evaluated measuring the PC headgroup order parameters from POPC:POPS (5:1)
mixtures\cite{roux90,NMRlipidsIV}.
Following our previous work \cite{NMRlipidsIV}, we evaluate the counterion binding affinity
to lipid bilayer in simulations containing PS using two different approaches.

@@

\begin{figure*}[!tbp] 
  \centering 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_nacl.pdf} 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_kcl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PC-PS_mix_PC} 
    The POPC head group order parameters and the P--N vector angle
    with respect to the membrane normal as a function of POPS content in a bilayer
    from ECC-lipid and Lipid17/Dang simulations with \ce{Na^+} (top) and \ce{K^+} (bottom) counterions.
    Experimental order parameter values are from Ref. \citenum{scherer87}
    and the signs from Ref. \citenum{ferreira16} (only \ce{Na^+} counterions, but shown also in bottom plots).
  }
\end{figure*}

\begin{figure*}[!tbp] 
  \centering 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPS_nacl.pdf} 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPS_kcl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PC-PS_mix_PS} 
    The POPS head group order parameters, the P--N and C$_\beta$--C$_\gamma$ vector angles (stars)
    with respect to the membrane normal as a function of POPC content in a bilayer
    from ECC-lipid and Lipid17/Dang simulations with \ce{Na^+} (top) and \ce{K^+} (bottom) counterions.
    Experimental order parameter values and signs for pure POPS are from Ref. \citenum{NMRlipidsIV} and values
    for POPC:POPS (5:1) mixture from Ref. \citenum{roux90} (only \ce{Na^+} counterions, but shown also in bottom plots).
    %The underestimated response of ECC-lipids with \ce{Na^+} counterions 
    %is likely due to a still slightly overestimated binding affinity of \ce{Na^+} to the phospholipids,
    %which is corroborated by the series with \ce{K^+} counterions (lower affinity than \ce{Na^+}),
    %where the response is closer to the experiment, which uses only \ce{Na^+} counterions. 
  }
\end{figure*}

@@

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPC_nacl.pdf} 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPS_nacl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PCPS} 
    Changes of the head group order parameters, and the angles of P--N and C$_\beta$--C$_\gamma$ (stars) vectors
    with respect to the membrane normal of POPC (left) and POPS (right) in a POPC:POPS (5:1) bilayer 
    as a function of \ce{NaCl} concentration from ECC-lipid and Lipid17/Dang simulations at 298 K.
    The y-axis for the $\alpha$-carbon results of POPS (middle right) is transferred
    with the same value for both order parameters such that the lower order
    parameter value from pure POPS is at zero to correctly illustrate the significant forking.
    Because experimental data with \ce{NaCl} is not available for POPC, the data for KCl and \ce{LiCl} (dashed line, left)
    are shown as lower and upper bounds for the response to \ce{NaCl}, respectively.
    The magnitudes of experimental order parameters are from Ref. \citenum{roux90} and signs from Ref. \citenum{ferreira16}.
    %which has a lower affinity to phospholipid bilayers compared to \ce{LiCl} \citep{roux90}. 
    %The orientation of the \ce{COO^-} group is defined as 
    %the connector from the $\beta$ carbon to the carbon in \ce{COO^-} (stars, bottom right). 
  }
  \end{figure}

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPC_kcl.pdf} 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPS_kcl.pdf} 
  \caption{\label{fig:delta_ordPar_KCl_PCPS}
    Changes of the head group order parameters, and the angles of P--N and C$_\beta$--C$_\gamma$ (stars) vectors 
    with respect to the membrane normal of POPC (left) and POPS (right) in a POPC:POPS (5:1) bilayer 
    as a function of \ce{KCl} concentration from ECC-lipid and Lipid17/Dang simulations 
    compared with experimental values from Ref. \citenum{roux90} (signs from Ref. \citenum{ferreira16}) at 298 K.
    The y-axis for the $\alpha$-carbon results of POPS (middle right) is transferred
    with the same value for both order parameters such that the lower order
    parameter value from pure POPS is at zero to correctly illustrate the significant forking.
  }
\end{figure}

@@

% This§: 
% adsorption and affinities of Na and K to PC:PS memb
% from densities and rel.surf.excess
% 
%in  Fig.~\ref{fig:POPS-counterions-dens} (only counterions)
%and Fig.~\ref{fig:nacl-dens_PCPS}        (additional salt concentrations), 
%
The influence of negatively charged PS on the counterion binding affinity can
be estimated by comparing the relative surface excesses with respect to water, $\Gamma ^{w} _{\rm ion}$,
between POPC:POPS (5:1) mixture and pure POPC \cite{melcr18}.
%Such a quantity compares the adsorption of ions to the adsorption of water molecules at an interface 
%without the necessity of defining a Gibbs dividing surface. \citep{melcr18, chattorajBOOK}
The value of $\Gamma ^{w} _{\rm Na}=-0.11 \pm 0.01 \mathrm{nm}^{-2}$ for 1 M sodium in pure ECC-POPC
simulation was reported in the previous work \cite{melcr18}.
The presence of $\sim$17\$ POPS in POPC bilayer increases the
%relative surface excess of sodium with respect to water from negative
value to $\Gamma ^{w} _{\rm Na}=0.092 \pm 0.005 \mathrm{nm}^{-2}$ 
for 0.621 M sodium, while value for potassium remains negative,
$\Gamma ^{w} _{\rm K}=-0.123 \pm 0.005 \mathrm{nm}^{-2}$ in POPC:POPS (5:1) mixture (Fig. \ref{fig:nacl-dens_PCPS}).
As discussed above, the sodium binding to POPC:POPS (5:1) mixture may be slightly overestimated also in
ECC-lipid simulations. The response of headgroup order parameters to the addition of NaCl are
slightly overestimated also in ECC-POPC simulations (Fig 3 in Ref. \citenum{melcr18}),
suggesting that similar small overestimation of sodium binding may also occur to a pure POPC bilayer.
Therefore, the overestimation may arise from the sodium interactions with PC headgroup and
increase in binding affinity due to PS lipids would not be affected.
%Interestingly enough, \ce{K^+} maintains negative values of $\Gamma^{w}_{K}$ even for the negatively charged PC:PS~(5:1) bilayer.
%This is also evident from the density plots in  Fig.~\ref{fig:nacl-dens_PCPS}
%showing that the density profile of \ce{K^+} along the membrane normal 
%decays clearly faster towards the membrane centre compared to the density of water. 
%In contrast, the density profile of \ce{Na^+} has a shoulder 
%at the region of the phosphate groups,
%which demonstrates that sodium cations adsorb weakly to the membrane interface compared to the aquaeous solution
%by yielding a small positive value of $\Gamma^{w}_{Na}$.

>>>>

\begin{table*}[tbp]
\centering
\caption{  List of MD simulations reporting their respective
           lipid composition,
           used counterions,
           added buffer concentrations,
           amounts of individual molecules,
           and the deployed models.
           The fatty acid chains for all lipids are plamitoyl (sn-1) and oleoyl (sn-2).
         }\label{tbl:sim-list}
\begin{tabular}{l l | r r r r | r r | c c }
   \multicolumn{2}{c}{ }  &  \multicolumn{4}{c}{added buffer conc. / mM}    & \multicolumn{2}{c}{no. molecules} &  \multicolumn{2}{c}{simulated with model}  \\
PC:PS ratio & counterions &  \ce{K^+}  &  \ce{Na^+} & \ce{Ca^{2+}} & \ce{Cl^-}      & \ce{H2O} &  PC:PS                 &  ECC-lipids  &  Lipid17    \\
  \hline
0:$\infty$ & \ce{K^+}   &      0  &      0  &      0  &      0  &  3600  &  0:72  &  X  &  X  \\ 
0:$\infty$ & \ce{Na^+}  &      0  &      0  &      0  &      0  &  3600  &  0:72  &  X  &  X  \\ 
  \hline
1:1 & \ce{K^+}   &      0  &      0  &      0  &      0  &  5253  &  64:64  &  X  &  X  \\ 
1:1 & \ce{Na^+}  &      0  &      0  &      0  &      0  &  5253  &  64:64  &  X  &  X  \\ 
  \hline
4:1 & \ce{K^+}   &      0  &      0  &      0  &      0  &  3600  &  48:12  &  X  &  -  \\ 
4:1 & \ce{Na^+}  &      0  &      0  &      0  &      0  &  3600  &  48:12  &  X  &  -  \\ 
  \hline
5:1 & \ce{K^+}   &      0  &      0  &      0  &      0  &  3600  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &      0  &      0  &      0  &      0  &  3600  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &      0  &      0  &     78  &     78  &  3561  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &      0  &      0  &    125  &    125  &  3561  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &      0  &      0  &    202  &    202  &  3561  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &      0  &      0  &    409  &    409  &  3522  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &      0  &      0  &    621  &    621  &  3483  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &      0  &    621  &      0  &    621  &  3483  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &      0  &   1510  &      0  &   1510  &  3377  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &      0  &   3002  &      0  &   3002  &  3213  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &    621  &      0  &      0  &    621  &  3483  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &   1510  &      0  &      0  &   1510  &  3377  &  60:12  &  X  &  X  \\ 
5:1 & \ce{Na^+}  &   3002  &      0  &      0  &   3002  &  3213  &  60:12  &  X  &  X  \\ 
  \hline
10:1 & \ce{K^+}  &      0  &      0  &      0  &      0  &  3600  &  60:6  &  X  &  -  \\ 
10:1 & \ce{Na^+} &      0  &      0  &      0  &      0  &  3600  &  60:6  &  X  &  -  \\ 
\end{tabular}
\end{table*}

@@

In the reference simulations with the standard Amber force field,
the Lipid14 parameters~\cite{dickson14} for POPC, the Lipid17 parameters for POPS \cite{lipid17-future},
the TIP3P water model~\cite{jorgensen83}, and ion models by Dang and coworkers~\cite{smith94,chang1999,dang2006} were used.
The previously generated \cite{botan15} Lipid14 parameters for POPC in Gromacs format were downloaded from Ref.~\citenum{lipid14files}. 
The Lipid17 parameters for POPS were obtained from AmberTools18 \cite{amber18} 
and converted to Gromacs format using acpype \cite{acpype}.  
The ion model by Dang and coworkers~\cite{smith94,chang1999,dang2006} was used with Amber lipids because
the default Amber ion parameters \cite{aqvist90} led to the artificial clustering of ions in solution \cite{NMRlipidsIV}.
In the ECC-lipids simulations,
the ECC-POPC parameters~\cite{melcr18} (available from Ref. \citenum{ECC-POPC_nacl_cacl2_files}), 
the ECC-POPS parameteres derived in this work (see above), 
the SPC/E~\cite{Berendsen1987} water model, and ECC-ions \cite{martinek17, kohagen16, Pluharova2014} were used.
The SPC/E water model was selected because its lower dielectric constant is consistent with the
ECC concept \cite{leontyev11,leontyev14}.
\begin{table}[tbp]
  \caption{Simulation parameters}
  \label{tbl:mdpar}
  \begin{tabular}{ll}
    simulation property & parameter   \\
    \hline
    time-step           & 2~fs         \\
    equilibration time  & 50~ns  \\
    total simulation time     & $\geq 1 \mu$s  \\
    temperature         & 298~K       \\
    thermostat          & v-rescale  \cite{bussi07}   \\
    barostat            & Parrinello-Rahman, semi-isotropic \cite{parrinello81} \\
    long-range electrostatics & PME  \cite{darden93}  \\
    cut-off scheme      & Verlet \cite{Pall13}      \\
    Coulomb and VdW cut-off & 1.0~nm \\
    constraints         & LINCS, only hydrogen atoms \cite{hess97} \\
    constraints for water & SETTLE  \cite{miyamoto92} \\
    \hline
  \end{tabular}
\end{table}

The C--H bond order parameters were calculated directly from Eq.~\ref{OPequation}.
The time average was first calculated for each lipid, and
the standard error of the mean from these individual values was then used as the
error estimate \cite{botan15,ollila16,NMRlipidsIV}.
Python program ({\tt scripts/calcOrderParameters.py}) that uses the
MDAnalysis library \cite{agrawal11,gowers16} is available in Ref.~\citenum{MATCHgit}. 
The ion number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.
The salt concentrations in buffer before solvating the lipids, reported in the used experimental data set \cite{roux90},
were calculated as $[{\rm salt}]=N_{\rm c} \times[{\rm water}]\,/\,N_{\rm w}$,
where $N_{\rm c}$ is the number of cations in simulation, $[{\rm water}]$\,=\,55.5~M and $N_{\rm w}$
is the number of water molecules in the simulation.
As discussed in our previous work \cite{NMRlipidsIV}, the hydration levels of
multilamellae are expected to be sufficiently similar in the used simulations and reference experiments \cite{roux90}.

@@

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[height=\figheight]{../img/ecc_pops/density_profiles_na-k-counterions_wat_phos_compar_purePOPS_ecclipids-lipid17.pdf}
  \includegraphics[height=\figheight]{../img/ecc_pops/density_profiles_na-k-counterions_wat_phos_compar_5PC-1PS_ecclipids-lipid17.pdf}
  \caption{\label{fig:POPS-counterions-dens}
    Number density profiles of \ce{K^{+}} and \ce{Na^{+}} counterions along the membrane normal axis
    in ECC-lipids (solid lines) and Lipid17/Dang (dashed lines) simulations of pure POPS (left) or POPC:POPS (5:1, right) bilayers.  
    The density profiles of phosphate groups and water are divided by 4 and 100, respectively.  
}
\end{figure} 
Binding affinity of monovalent ions to lipid bilayers depends strongly on
force field parameters \cite{catte16,NMRlipidsIV},
which is the case also here as the sodium counterion binding affinity to
POPS lipid bilayer is stronger in Lipid17/ Dang simulation than in
ECC-POPS simulation (Fig. \ref{fig:POPS-counterions-dens}).
The cation binding affinities to lipid bilayers in simulations have been previously evaluated
against experiments using the order parameters of $\alpha$ and $\beta$ carbons in the
phosphatidylcholine (PC) lipid headgroup (see Fig.~\ref{fig:chemstruct_pc_ps} for the labeling) \cite{catte16,melcr18,NMRlipidsIV},
which decrease proportionally to the bound charge according to the ''electrometer concept'' \citep{seelig87}.
Althought the response of PS lipid headgroup order parameters to the bound charge is also systematic,
it is less well undestood and the addition of cations may cause phase transitions in negatively charged
bilayers \cite{feigenson86,mattai89,roux91,roux90}. Therefore, the cation binding affinity to bilayers with
PS lipids has been evaluated measuring the PC headgroup order parameters from POPC:POPS (5:1)
mixtures\cite{roux90,NMRlipidsIV}.
Following our previous work \cite{NMRlipidsIV}, we evaluate the counterion binding affinity
to lipid bilayer in simulations containing PS using two different approaches.

@@

\begin{figure*}[!tbp] 
  \centering 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_nacl.pdf} 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_kcl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PC-PS_mix_PC} 
    The POPC head group order parameters and the P--N vector angle
    with respect to the membrane normal as a function of POPS content in a bilayer
    from ECC-lipids and Lipid17/Dang simulations with \ce{Na^+} (top) and \ce{K^+} (bottom) counterions.
    Experimental order parameter values are from Ref. \citenum{scherer87}
    and the signs from Ref. \citenum{ferreira16} (only \ce{Na^+} counterions, but shown also in bottom plots).
  }
\end{figure*}

\begin{figure*}[!tbp] 
  \centering 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPS_nacl.pdf} 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPS_kcl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PC-PS_mix_PS} 
    The POPS head group order parameters, the P--N and C$_\beta$--C$_\gamma$ vector angles (stars)
    with respect to the membrane normal as a function of POPC content in a bilayer
    from ECC-lipids and Lipid17/Dang simulations with \ce{Na^+} (top) and \ce{K^+} (bottom) counterions.
    Experimental order parameter values and signs for pure POPS are from Ref. \citenum{NMRlipidsIV} and values
    for POPC:POPS (5:1) mixture from Ref. \citenum{roux90} (only \ce{Na^+} counterions, but shown also in bottom plots).
    %The underestimated response of ECC-lipids with \ce{Na^+} counterions 
    %is likely due to a still slightly overestimated binding affinity of \ce{Na^+} to the phospholipids,
    %which is corroborated by the series with \ce{K^+} counterions (lower affinity than \ce{Na^+}),
    %where the response is closer to the experiment, which uses only \ce{Na^+} counterions. 
  }
\end{figure*}

@@

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPC_nacl.pdf} 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPS_nacl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PCPS} 
    Changes of the head group order parameters, and the angles of P--N and C$_\beta$--C$_\gamma$ (stars) vectors
    with respect to the membrane normal of POPC (left) and POPS (right) in a POPC:POPS (5:1) bilayer 
    as a function of \ce{NaCl} concentration from ECC-lipids and Lipid17/Dang simulations at 298 K.
    The y-axis for the $\alpha$-carbon results of POPS (middle right) is transferred
    with the same value for both order parameters such that the lower order
    parameter value from pure POPS is at zero to correctly illustrate the significant forking.
    Because experimental data with \ce{NaCl} is not available for POPC, the data for KCl and \ce{LiCl} (dashed line, left)
    are shown as lower and upper bounds for the response to \ce{NaCl}, respectively.
    The magnitudes of experimental order parameters are from Ref. \citenum{roux90} and signs from Ref. \citenum{ferreira16}.
    %which has a lower affinity to phospholipid bilayers compared to \ce{LiCl} \citep{roux90}. 
    %The orientation of the \ce{COO^-} group is defined as 
    %the connector from the $\beta$ carbon to the carbon in \ce{COO^-} (stars, bottom right). 
  }
  \end{figure}

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPC_kcl.pdf} 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPS_kcl.pdf} 
  \caption{\label{fig:delta_ordPar_KCl_PCPS}
    Changes of the head group order parameters, and the angles of P--N and C$_\beta$--C$_\gamma$ (stars) vectors 
    with respect to the membrane normal of POPC (left) and POPS (right) in a POPC:POPS (5:1) bilayer 
    as a function of \ce{KCl} concentration from ECC-lipids and Lipid17/Dang simulations 
    compared with experimental values from Ref. \citenum{roux90} (signs from Ref. \citenum{ferreira16}) at 298 K.
    The y-axis for the $\alpha$-carbon results of POPS (middle right) is transferred
    with the same value for both order parameters such that the lower order
    parameter value from pure POPS is at zero to correctly illustrate the significant forking.
  }
\end{figure}

@@

% This§: 
% adsorption and affinities of Na and K to PC:PS memb
% from densities and rel.surf.excess
% 
%in  Fig.~\ref{fig:POPS-counterions-dens} (only counterions)
%and Fig.~\ref{fig:nacl-dens_PCPS}        (additional salt concentrations), 
%
The influence of negatively charged PS on the counterion binding affinity can
be estimated by comparing the relative surface excesses with respect to water, $\Gamma ^{w} _{\rm ion}$,
between POPC:POPS (5:1) mixture and pure POPC \cite{melcr18}.
%Such a quantity compares the adsorption of ions to the adsorption of water molecules at an interface 
%without the necessity of defining a Gibbs dividing surface. \citep{melcr18, chattorajBOOK}
The value of $\Gamma ^{w} _{\rm Na}=-0.11 \pm 0.01 \mathrm{nm}^{-2}$ for 1 M sodium in pure ECC-POPC
simulation was reported in the previous work \cite{melcr18}.
The presence of $\sim$17\$ POPS in POPC bilayer increases the
%relative surface excess of sodium with respect to water from negative
value to $\Gamma ^{w} _{\rm Na}=0.092 \pm 0.005 \mathrm{nm}^{-2}$ 
for 0.621 M sodium, while value for potassium remains negative,
$\Gamma ^{w} _{\rm K}=-0.123 \pm 0.005 \mathrm{nm}^{-2}$ in POPC:POPS (5:1) mixture (Fig. \ref{fig:nacl-dens_PCPS}).
As discussed above, the sodium binding to POPC:POPS (5:1) mixture may be slightly overestimated also in
ECC-lipids simulations. The response of headgroup order parameters to the addition of NaCl are
slightly overestimated also in ECC-POPC simulations (Fig 3 in Ref. \citenum{melcr18}),
suggesting that similar small overestimation of sodium binding may also occur to a pure POPC bilayer.
Therefore, the overestimation may arise from the sodium interactions with PC headgroup and
increase in binding affinity due to PS lipids would not be affected.
%Interestingly enough, \ce{K^+} maintains negative values of $\Gamma^{w}_{K}$ even for the negatively charged PC:PS~(5:1) bilayer.
%This is also evident from the density plots in  Fig.~\ref{fig:nacl-dens_PCPS}
%showing that the density profile of \ce{K^+} along the membrane normal 
%decays clearly faster towards the membrane centre compared to the density of water. 
%In contrast, the density profile of \ce{Na^+} has a shoulder 
%at the region of the phosphate groups,
%which demonstrates that sodium cations adsorb weakly to the membrane interface compared to the aquaeous solution
%by yielding a small positive value of $\Gamma^{w}_{Na}$.
